Comparing vector-based and ACT-R memory models using large-scale datasets:\protect\\User-customized hashtag and tag prediction on Twitter and StackOverflow

### 2014-11-17 writeups/psycMethods/psycMethods.tex
foo

\documentclass[man,floatsintext,donotrepeattitle]{apa6}

@@

\title{Comparing vector-based and ACT-R memory models using large-scale datasets:\protect\\User-customized hashtag and tag prediction on Twitter and StackOverflow}
%\shorttitle{Comparing memory models of tag retrieval}
\shorttitle{}

\author{Clayton Stanley}
\affiliation{Rice University}

@@

\def\committee{
  Michael D. Byrne, Chair \\
  Professor, Psychology and Computer Science \and
  Phillip Kortum \\
  Assistant Professor, Psychology \and
  David Lane \\
  Associate Professor, Psychology, Statistics, and Management \and
  Devika Subramanian \\
  Professor, Computer Science and Electrical Engineering \\
}

% Replace the title page in the apa6 style sheet with the title page from the Rice template
% The title page style was taken from Rice University LaTeX PhD template: http://scholarship.rice.edu/handle/1911/21747
% The following code for the abstract page was taken from the apa6 \maketitle command.
\makeatletter
\def\makeDisCover{
  \def\maketitle{
    \begin{titlepage}
      \singlespacing
      \null
      \vfill
      \vbox to 0pt{\vss\vbox to 8.75in{\parskip 0pt \parindent 0pt\centering
	{\uppercase\expandafter{Rice University} \par} \vskip 0pt plus 3fil
	{\bf \@title \par} \vskip 0pt plus 1.3fil
	{by \par} \vskip 0pt plus 1fil
	{\bf \@author \par} \vskip 0pt plus 3fil
	{\sc A Thesis Submitted \\
	  in Partial Fulfillment of the \\
	Requirements for the Degree \par} \vskip 0pt plus 1.1fil
	{\bf Doctor of Philosophy \par} \vskip 0pt plus 4.5fil
	\begingroup
	\leftskip .45\textwidth \rightskip 0pt plus 1fil
	{\sc Approved, Thesis Committee: \par} \vskip 2.5em
	\moveright\leftskip\vbox{\hrule width \leftskip}
	\def\and{\par \vskip 2.25em minus 1em\moveright\leftskip\vbox{\hrule width \leftskip}}
	\committee
	\par \endgroup
	\vskip 0pt plus 2fil
	{Houston, TEXAS \par} \vskip 0pt 
	{December \number\year \par}
      }}
      \doublespacing
      \newpage
  %BDB\hyphenpenalty 10000
      \fussy
      \@ifundefined{@abstract}{}{%
	\thispagestyle{empty}
	\section{\normalfont\normalsize \abstractname}% BDB
	\begingroup
	\parskip \z@ \parindent \z@ \par
	\null \vskip .1in \centering
	\@title
	\vskip .3in
	{by \par}
	\@author \par
	\vskip .5in
      \endgroup
      \@abstract\par% BDB
      \@ifundefined{@keywords}{}{%
	\setlength{\parindent}{0.4in}% BDB
	\indent\textit{\keywordname:} \@keywords%
      }%
      \thispagestyle{empty}
      \newpage
    }
    \@ifundefined{def@donotrepeattitle}{
      \section{\protect\normalfont{\@title}}
    }{}%
    \raggedright%
  \end{titlepage}
  \setlength{\parindent}{0.4in}%
}}
\makeatother

\makeDisCover

%\keywords{}

@@

\pagenumbering{roman}

@@

\note{\today}

@@

\newpage

\pagenumbering{arabic}

@@

The core question motivating this research is:
What kinds of cognitively-plausible models can best predict the tags that a user on a social media site uses?
This question can be tested by generating a prediction for the most likely used hashtag whenever a user is about to generate a hashtag, and then comparing the user's chosen hashtag to the model's prediction.
This process can be framed as a memory retrieval problem.
The process of suggesting a relevant hashtag to a user can be thought of as a memory retrieval request for a hashtag, given prior hashtag use and current context.
That is, each user on a social media site has created content that contains a set of tags within the context of each post.

Co-occurrence-based modeling has been shown as a potentially useful approach when predicting Twitter hashtag use \parencite{Efron2010} and StackOverflow tag use \parencites{Kuo2011,Stanley2013}.
Several memory retrieval models are based on this broad methodology, where a count is maintained of the times each contextual element (such as a word in a post) co-occur with a tag.
At least two of the current memory retrieval models are based on a co-occurrence methodology:
ACT-R's declarative memory (DM) retrieval theory and random permutation vector-based memory systems.

@@

\subsubsection{Comparison Between ACT-R and Vector-Based Memory Systems}

ACT-R's declarative memory and vector-based memory systems are substantially different in their mathematical structure.
Nonetheless, both can successfully model classic behavioral patterns found in word pairing experiments (e.g., \cite{Rutledge2008} modeling the ``fan effect'').

An advantage of vector-based models is that word order can be easily incorporated into the single co-occurrence representation \parencite{Jones2007}.
Word order information on sites like Twitter may contain highly predictive pieces of information, as it is likely that specific words immediately precede specific hashtags.
Word order has been incorporated into ACT-R sentence-parsing models previously, but this has not been done using declarative memory.
Rather, production rules were defined that ensured that the model followed a set of rules (e.g., left-to-right reading) when parsing sentences \parencite{Lewis2006}.
It is certainly true that at least some production rules are needed when choosing words and creating content. 
However, it is unexplored if some of the process of sentence construction and word selection can be represented in the ACT-R declarative memory system.

ACT-R's declarative memory system places strong constraints on how a user's prior knowledge and experience influences the likelihood that a particular memory item is retrieved \parencite{Anderson2007}.
This decay rate equation formalizes how recency and frequency of memory use relate to the prior probability that a memory will be retrieved.
However, it is unexplored how a user's prior knowledge should influence retrieval for vector-based models.

@@

Latent Semantic Analysis \parencite{Landauer1997} (LSA) is an alternative representation of declarative memory.
Both ACT-R's Bayesian declarative memory model and LSA compute activations based on a co-occurrence matrix.
However, LSA performs further processing on that matrix in order to collapse words that have similar meaning.
The process starts with a set of N documents.
A word by document co-occurrence matrix is built by counting the number of times each word appears in each document.
The dimensionality of the full word frequency co-occurrence matrix is then reduced, while maintaining as much of the variance in the original full matrix as possible.
LSA uses the Singular Value Decomposition process to reduce the dimensionality of the word by document co-occurrence matrix.
The strength of association between two words is measured by the cosine of the two word vectors across the reduced dimension space.

@@

One of the strengths of vector-based systems is that word order can naturally be represented alongside word co-occurrence in these distributed representations.
Representing word order is a case where both the encoding and decoding operations are applied.
As a motivating example, say that one wants to encode the first and second words immediately preceding a hashtag in a tweet.
Simply counting co-occurrences of the two immediately preceding words and hashtags does not include the order information of the words in the representation.
With this representation, it would be impossible to tell the difference between 1-back words that immediately preceded the hashtag, and 2-back words that occurred just prior to these 1-back words.
Order information is lost if a simply co-occurrence counting technique is used.

With vector-based systems, order information is included by creating an environment vector ($e_{bind(i,\Phi)}$) that is a function of environment vectors for both the word ($e_{i}$) and location ($\Phi$).
This function has the useful property that the resulting environment vector and the location vector can be inverted to return the original word's environment vector.
Circular convolution and inverse circular convolution (i.e., circular correlation) are used by \textcite{Plate1995} and \textcite{Jones2007} to implement these operations.
This allows for retrieval requests such as ``playing \#$\Phi$,'' ``just read \#$\Phi$,'' and ``vector-based $\Phi$ systems.''

\subsubsection{BEAGLE}

\textcite{Jones2007} created the BEAGLE vector-based memory representation.
Each memory vector ($m_{i}$) is a summed representation of unordered and ordered word co-occurrences ($c_{i} + o_{i}$).
This model uses circular convolution and circular correlation as a way to encode and decode word order.
BEAGLE performed similar to LSA when trained on the TASA and tested on the TOEFL (\num{55.6}\% compared to \num{55.3}\%).
Further, it was shown that incorporating word order information into the same representation resulted in minimal data loss (less than 1 percent of total predictive variance).
This model provided an efficient algorithm for using a single representation for unordered and ordered information that showed performance similar to if not better than LSA.

@@

In order to represent word order information, a function must be used that converts an environment vector for a set of words and positions (e.g., ``stack $\Phi$'') into a new uncorrelated environment vector.
The BEAGLE model uses circular convolution and circular correlation as the encoding and decoding operations.
However, that is certainly not the only operation that can generate new uncorrelated vectors from a set of memory vectors and positions.
\textcite{Sahlgren2008} used a much simpler random permutation method for this operation.

@@

For example, suppose that the words ``stack overflow'' appear in a sentence, and I want to represent that ``stack'' preceded the word ``overflow.''
The ``stack'' environment vector ($e_{stack}$) is shifted one to the left, producing an environment vector for ``stack'' preceding a word ($e_{stack^{-1}}$).
That environment vector is then added to the memory vector for ``overflow'' ($m_{overflow} = m_{overflow'} + e_{stack^{-1}}$), so that ``overflow'' remembers that ``stack'' has preceded it in a sentence.

A resonance retrieval for a word following ``stack'' (``stack $\Phi$'') is done by creating a memory vector out of all information in context.
In this case, assume only a single piece of order information is available, that ``stack'' precedes the word to retrieve.
So the memory vector for context is the environment vector for one-back ``stack'' ($m_{C} = e_{stack^{-1}}$).
The memory vector for ``overflow'' ($m_{overflow}$) will be correlated the highest with this context vector ($m_{C}$) since ``overflow'' contains this vector.

\subsubsection{Comparison between BEAGLE and Random Permutations}

Random permutations uses one of simplest operations possible to create a new environment vector.
This operation is much less costly than circular convolution used by the BEAGLE model, which allows random permutations to scale better and handle larger datasets.
Further, even though random permutations is a simpler representation than BEAGLE, it also performs better when trained on the same sized corpus.

\textcite{Recchia2010} found that random permutations could use smaller vectors than circular convolution approaches (e.g., BEAGLE) to produce equivalent performance (better compression),
and also scale to larger datasets (better computational efficiency).
These results make a strong case to favor random permutations over circular convolution when implementing vector-based memory representations.

@@

Both random permutations and LSA calculate word similarity on a reduced matrix.
For LSA, the original word by document matrix is compressed to a factor by document matrix and the number of factors is much less than the number of words.
For random permutations, each tag is represented by a compressed vector with a length much less than the number of words.

Information is lost with this compression in both cases: by starting with a compressed representation with random permutations or by reducing to a compressed representation with LSA.
However, the way that the compression works in these two cases is not the same.
LSA uses SVD to find words that have highly correlated tag occurrence distributions (i.e., latent concepts), and then pools those co-occurrence counts into a single dimension.
A vector-based approach like BEAGLE or random permutations essentially randomizes the mapping of the words on the reduced set of dimensions, and does not take into account word similarity when generating the mapping.
Nonetheless, random permutations have been shown to perform similar to if not better than LSA when trained on the TASA and tested on the TOEFL \parencites{Sahlgren2008,Jones2007}.

@@

\textcite{Kuo2011} was the first to model tag use for StackOverflow posts.
His set of models were originally designed for next-word prediction for large text document collections.
\citeauthor{Kuo2011} examined the accuracy of three different commonly-used models for next-word prediction: a Bayesian model, a K-nearest-neighbors model, and a Latent Semantic Analysis (LSA) model.
In order to structure the models to work for StackOverflow tag prediction, the models would request the most-likely next word after a user created a post, and restrict the space of possible words to only tags.
He tested the set of next-word prediction models to generate predictions for the most likely tags that a user would choose when asking a question on the site.
Interestingly, the Bayesian co-occurrence model outperformed the more complicated and computationally expensive LSA and K-nearest-neighbors models.
This Bayesian model can accurately predict a user-chosen tag at 47\% accuracy.

% ref: http://tex.stackexchange.com/questions/53513/hyperref-token-not-allowed
\paragraph{\texorpdfstring{\textcite{Stanley2013}}{Stanley2013}'s ACT-R DM Model for StackOverflow}

\textcite{Stanley2013} also modeled a user's chosen tags for a post on the StackOverflow site.
They modified the standard ACT-R model in order to more accurately predict tag use on StackOverflow.
A formal description of the modified ACT-R Bayesian model is included in Table \ref{tabModACTRModel}.

\begin{table}[!ht]
  \caption{\citeauthor{Stanley2013}'s StackOverflow tag prediction model}
  \label{tabModACTRModel}
  {\tabulinesep=1.2mm
    \begin{tabu}{ll}
      \hline
      Common Name &  Equation \\
      \hline
      Activation & 		$A_{i} = B_{i} + \sum_{j\in T}^{ } W_{j} S_{ji} + \sum_{j\in B}^{ } W_{j} S_{ji} - O$ \\
      Base Level & 		$B_{i} = log \frac{p_{i}}{1-p_{i}}$ \\
      Strength Assoc. &		$S_{ji} = log \frac{p(i|j)}{p(i))} = log \frac{NN_{ji}}{N_{Row}(j)N_{Col}(i)}$ \\
      Co-occurrence &		$N_{ji} = \sum_{}^{}{observed_{ji}}$ \\
      Observations &		$N = \sum_{}^{}{\sum_{}^{}{N_{ji}}}$ \\
      Attentional Weight& 	$W_{j} = W \frac{E_{j}} {\sum_{}^{} {E_{j}}} $ \\
      Scaled Entropy & 		$E_{j} = 1-\frac{H_{j}}{H_{max}}$ \\
      Entropy & 		$H_{j} = -\sum_{i=1}^{N}p(i|j)log\left (  p(i|j) \right )$ \\
      Offset & 			$O = \frac{1}{5}\sum_{i\in top 5}^{ } A_{i}$ \\
      \hline
    \end{tabu}
  }
\end{table}

The activation of each tag ($A_{i}$) is a function of four components:
That tag's base-level activation ($B_{i}$), contextual activation for the words in the title, activation for words in the body, and an offset term ($O$).
This model has been modified from the standard ACT-R model in three ways:
% FIXME: four ways; also base level term
Two separate contextual components were used (words in title and body of post), an offset term was added, and an entropy measure was used to compute the attentional weight ($W_{j}$).

The offset term was added to normalize the mean activation across retrievals.
This allowed the use of a logistic regression statistical technique to calibrate the weights for each of the terms.
The title and body contextual components were separated since it seemed plausible that the words in the title would be better predictors of the tag used than the words in the body.
It turned out that the words in the body were a better predictor of tag use \parencite{Stanley2013}, most likely because there are many more words in the body than the title of the post.

The scaled entropy measure $E_{j}$ was chosen for the attentional weight ($W_{j}$) term because it has a strong theoretical foundation and it predicts well \parencite{Dumais1991}.
It is based on Shannon's Information Theory, which measures the amount of information content (predictiveness) of a word.
So the scaled entropy ($E_{j}$) of a context word can be thought of as the amount of predictiveness that that word has to \emph{any} tag.
This means that a stop word like ``the'' will have a scaled entropy measure close to 0, while a highly-predictive word like \emph{macro} will have a scaled entropy measure close to 1.
Looking at $W_{j}$, if $E_{j}$ is low then $W_{j}$ is low, meaning that the scaled entropy measure helps allocate limited attentional resources to predictive contextual elements.

@@

Several hashtag recommendation models for Twitter have been developed recently.
One of the first models used a Bayesian co-occurrence statistical technique to predict the most likely hashtag associated with a post as a function of prior hashtag use and context \parencite{Mazzia2009}.
The model presented by \textcite{Mazzia2009} is quite similar to an ACT-R declarative retrieval model.
The main potential difference is that --- like \textcite{Kuo2011} and \textcite{Stanley2013} --- the global prior likelihood of a hashtag is computed without taking into account the specific user's past history.
Instead, the global prior is an overall average frequency of hashtag use across all users.
Further, no recency information was used to compute the global prior, so hashtags used in the recent past were not weighted more heavily. 
This was most likely not much of an issue for \citeauthor{Mazzia2009}'s study, since all hashtags were collected within a relatively short period of time (about 3 months).

Several other models have taken a tweet-centered approach to hashtag prediction, where suggested hashtags are collected from hashtags used from similar tweets.
\textcites{Li2011, Zangerle2011, Kywe2012} all store a content vector for each tweet, and then compute a word co-occurrence-based similarity score between a composed tweet and the rest of the tweets in the database.
Each method collects the hashtags used from a set of the most similar tweets, then ranks the hashtags in the set, and presents the top 5-10 hashtags to the user.

\subsubsection{State of Research on StackOverflow and Twitter}

\textcite{Stanley2013} used an ACT-R based Bayesian memory model to predict human tagging on the StackOverflow site, but to our knowledge no vector-based memory model has been tested on the site.
Both types of models have been successfully incorporated as the declarative memory component of the ACT-R cognitive architecture \parencite{Rutledge2007},
so it should be straightforward to test a vector-based model on the StackOverflow site.

Also, a vector-based retrieval model has not yet been tested on Twitter.
One of the advantages of vector-based models is that word order can be easily incorporated into the single co-occurrence representation \parencite{Jones2007}.
It seems quite likely that words used just before a hashtag in a tweet (combined with their position) will be highly predictive of the hashtag chosen.
To our knowledge, word order has not yet been taken into account for any of the Twitter hashtag-prediction models that have already been tested.

@@

One of the main purposes of this research is to improve on these user-agnostic statistical models and include each specific user's past tagging history as a model component.
Using customized priors that are tailored to a user's specific tagging history should be crucial for sites like Twitter where the possible tag space is infinite,
but should also improve performance for more constrained sites like StackOverflow since each user is not interested in all of the possible tags on the site.

@@

Two models were evaluated for hashtag prediction on both the Twitter and StackOverflow datasets: a vector-based model and Bayesian retrieval model.
The random permutation model \parencite{Sahlgren2008} in Table \ref{tabRandPermModel} was used for the vector-based model, since it has shown to perform better than the BEAGLE model \parencite{Recchia2010}.
The modified ACT-R Bayesian model \parencite{Stanley2013} in Table \ref{tabModACTRModel} was used for the Bayesian model, since it has already been applied to tag use on StackOverflow posts.

@@

An additional popular-hashtags dataset was collected for Twitter.
The purpose of this dataset is to test the accuracy of each model's predicted most likely hashtag(s) for a tweet.
The dataset contains approximately three million tweets where at least one of the 400 most popular hashtags are used.
A quota of three million tweets was selected because previously research testing Bayesian co-occurrence models on large-scale datasets has used on the order of one million documents
\parencites{Stanley2013,Douglass2010,Budiu2007,Kuo2011}.
Tweets were collected in real time until the tweet quota was reached.

@@

A custom tokenization algorithm created by \textcite{Owoputi2013} was used to chunk the words in the text.
This algorithm was specifically created for tokenizing Twitter content.
However, during testing it was also found to be robust and well suited for StackOverflow content.
The initial model results with the new tokenizer were qualitatively similar to the model results found in \textcite{Stanley2013} where the Python Natural Language Processing (NLP) toolkit \parencite{Bird2009} was used.
This is most likely because the language for the samples taken for both sites is informal English,
and it seams reasonable to think that tokenizing all of the informal language and characters included in Twitter is a much more difficult task than tokenizing most of the words used on StackOverflow.
So the same tokenization algorithm was used for both StackOverflow and Twitter, since performance was similar to \textcite{Stanley2013} for StackOverflow,
and it seemed beneficial to use the same algorithm for both datasets to make comparisons across datasets more equitable.

@@

The decay rate parameter ($d$) represents how fast the activation for an individual presentation of a chunk decays from declarative memory.
Higher values mean that chunk activation decays faster, which means that the most recently presented chunks will have the highest activation, and the total number of times that a chunk has been presented matters less.
Lower decay rate values flip the importance of frequency and recency:
Lower values mean that chunks that have occurred most frequently will have the highest activation,
and activation depends less on the specific chunks that have been observed in the recent past.

@@

Both the standard form and optimized learning form of $B_{i}$ were used to predict each user's specific tag history.
To do this, a user's entire tag history was collected, and then each tag instance from oldest to newest was examined.
For each of these tagging instances, the model generated an activation value for each possible tag based on that tag's frequency and recency of use. 
This means, for example, that the model always missed characterizing the first observed tag for each user, since at that point the model does not have any history of tag use for that user.
This also means that the tag space for each user is constrained and limited to the tags that each user has used in the past.

@@

\subsubsection{Best-fit Decay Rate Values}

To identify the best decay rate value for the dataset slices,
each user's best decay rate value was taken and the results are plotted in histogram form in Figures \ref{figPriorSOQHistDsOL} and \ref{figPriorTwitterHistDsOL} respectively.
The best-fit values are close to the \num{0.5} default value set in ACT-R when the optimized learning form is used.
Comparing across optimized learning and the standard form, the best-fit values for the standard form are slightly higher for both the StackOverflow and Twitter dataset slices.
Comparing best-fit values will be more thoroughly tested when looking at the values across all dataset slices.

\begin{figure}[!htbp]
  {%
    \setlength{\fboxsep}{0pt}%
    \setlength{\fboxrule}{1pt}%
    \hfill
    \subfigure[][Standard form]{\includegraphics[width=8cm]{visHistD-SOQgt500r2-topHashtagPostPriorStd-crop.pdf}}
    \hfill
    \subfigure[][Optimized learning form]{\includegraphics[width=8cm]{visHistD-SOQgt500r2-topHashtagPostPriorOL2-crop.pdf}}
    \caption{
      Best-fit decay rate for a single dataset slice for StackOverflow.
      The histogram represents the count of users for each best-fit decay rate value.
      Each bar represents a discrete decay rate value that was explored.
  }
    \label{figPriorSOQHistDsOL}
  }%
\end{figure}

\begin{figure}[!htbp]
  {%
    \setlength{\fboxsep}{0pt}%
    \setlength{\fboxrule}{1pt}%
    \hfill
    \subfigure[][Standard form]{\includegraphics[width=8cm]{visHistD-TFollowgt10Mr2-topHashtagPostPriorStd-crop.pdf}}
    \hfill
    \subfigure[][Optimized learning form]{\includegraphics[width=8cm]{visHistD-TFollowgt10Mr2-topHashtagPostPriorOL2-crop.pdf}}
    \caption{
      Best-fit decay rate for a single dataset slice for Twitter.
      Best-fit values were slightly higher for Twitter, so the range of the histogram is higher for Twitter than StackOverflow.
  }
    \label{figPriorTwitterHistDsOL}
  }%
\end{figure}

@@

The decay rate parameter in $B_{i}$ determines how fast a presentation of a chunk in the past decays over time.
Higher values of $d$ mean that activation decays more quickly, which means that the chunks used more recently are the most active and the total frequency count of a chunk matters less.
Looking back at Figure \ref{figPriorDecay}, the optimal decay rate parameter when the standard form of $B_{i}$ is used is higher for Twitter than StackOverflow.
It makes sense that the lifetime of a specific hashtag on Twitter is shorter overall than the lifetime of a tag (e.g., \emph{C\#}) on StackOverflow,
since hashtag topics are generally more transient on Twitter than StackOverflow (e.g., hashtags for specific political events or upcoming conferences).
This means that the decay rate for Twitter should be higher than StackOverflow, which is what is observed.

@@

Chapter 5 will examine how model accuracy is influenced by two methodological concerns (adding an offset term and examining the effect of dataset subset for Twitter)
and four architectural concerns
(method of handling stop words, influence of dataset size, method to combine prior and context, and word order).
The two methodological concerns were examined to ensure that the technique to calibrate model parameters was accurate and also that results for Twitter were not specific to the subset of data collected.

The four architectural concerns were examined for the following reasons:
During exploratory evaluation, model accuracy seemed to be highly sensitive to the method of dealing with the commonly-used stop words, so several different methods were examined in more detail.

Due to research from \textcite{Budiu2007} and \textcite{Recchia2010}, model accuracy for the Bayesian and vector-based models is influenced by the number of documents in the dataset,
so the effect of dataset size on accuracy was examined in more detail.
Specifically,
since both the Bayesian and random-permutation models are being evaluated, analyzing the effect of dataset size on accuracy in this research will allow direct comparisons between the two models on a common dataset.
This has not yet been possible since the two models have never been compared on the same dataset within the same study.

One of the primary goals of this research is to identify the best method to combine the model term for past user behavior with the context term for the random-permutation model,
so two different ways to add the model terms were tested.

Finally, previous research has shown that including word order in the vector-based models improves performance \parencites{Jones2007, Sahlgren2008},
so word order was also examined to see if the effect could be replicated and also to measure the overall importance of word order on model accuracy relative to other architectural manipulations.

@@

In order to best combine these predictors, a logistic regression statistical technique similar to \textcite{Stanley2013} was used to determine the most optimal weights for each term.
\textcite{Stanley2013} weighted each term in the total activation equation (Table \ref{tabModACTRModel}) to maximize model accuracy.
\citeauthor{Stanley2013} found that the optimal weights for the StackOverflow model for the prior term, title context term, and body context term were \num{.62}, \num{.93}, and \num{1.75} respectively.

The optimal weights for each model component were computed in the same manner as \textcite{Stanley2013}:
When a user creates a tag for a post, a retrieval request for the model is made.
For each request, the model returns a set of tags (all tags that have been observed for that user in the past) and activations associated with each tag.
An activation value is computed for each model component (e.g., prior tag use, context) for each tag.
These values are recorded, and tag instances that match what the author actually tagged are marked as a 1, while all others are marked with a 0.
This process is repeated across all posts where a user chooses a tag, and across all users in the dataset.
Once all of the results are aggregated, a logistic regression is run where the optimal weights for each model component are found that maximize the model's ability to correctly label tags as 0 or 1 based on total activation.

>>>>

%\documentclass[man,floatsintext,donotrepeattitle]{apa6}
\documentclass[man,donotrepeattitle]{apa6}

@@

\title{Comparing vector-based and ACT-R memory models using large-scale datasets:\protect\\User-customized hashtag and tag prediction on Twitter and StackOverflow}
\shorttitle{Comparing memory models of tag retrieval}
%\shorttitle{}

\author{Clayton Stanley and Michael D. Byrne}
\affiliation{Rice University}

@@

\keywords{ACT-R declarative memory theory, vector-based models, LSA, machine learning}

@@

@@

%\note{\today}

@@

%\newpage

@@

The core question motivating this research is:
What kinds of cognitively-plausible models can best predict the tags that a user on a social media site uses?
This question can be tested by generating a prediction for the most likely used hashtag whenever a user is about to generate a hashtag, and then comparing the user's chosen hashtag to the model's prediction.
This process can be framed as a memory retrieval problem.
The process of suggesting a relevant hashtag to a user can be thought of as a memory retrieval request for a hashtag, given prior hashtag use and current context.

Co-occurrence-based modeling has been shown as a potentially useful approach when predicting Twitter hashtag use \parencite{Efron2010} and StackOverflow tag use \parencites{Stanley2013}.
Several memory retrieval models are based on this broad methodology, where a count is maintained of the times each contextual element (such as a word in a post) co-occur with a tag.
At least two of the current memory retrieval models are based on a co-occurrence methodology:
ACT-R's declarative memory (DM) retrieval theory and random permutation vector-based memory systems.

@@

@@

Latent Semantic Analysis \parencite{Landauer1997} (LSA) is an alternative representation of declarative memory.
Both ACT-R's Bayesian declarative memory model and LSA compute activations based on a co-occurrence matrix.
However, LSA performs further processing on that matrix in order to collapse words that have similar meaning.
LSA uses the Singular Value Decomposition process to reduce the dimensionality of the word by document co-occurrence matrix, while maintaining as much of the variance in the original full matrix as possible.
The strength of association between two words is measured by the cosine of the two word vectors across the reduced dimension space.

@@

One of the strengths of vector-based systems is that word order can naturally be represented alongside word co-occurrence in these distributed representations.
With vector-based systems, order information is included by creating an environment vector ($e_{bind(i,\Phi)}$) that is a function of environment vectors for both the word ($e_{i}$) and location ($\Phi$).
This function has the useful property that the resulting environment vector and the location vector can be inverted to return the original word's environment vector.
Circular convolution and inverse circular convolution (i.e., circular correlation) are used by \textcite{Plate1995} and \textcite{Jones2007}'s BEAGLE model to implement these operations.
This allows for retrieval requests such as ``playing \#$\Phi$,'' ``just read \#$\Phi$,'' and ``vector-based $\Phi$ systems.''

@@

In order to represent word order information, a function must be used that converts an environment vector for a set of words and positions (e.g., ``stack $\Phi$'') into a new uncorrelated environment vector.
\textcite{Sahlgren2008} used a simple random permutation method for this operation.

@@

@@

Both vector-based models and LSA calculate word similarity on a reduced matrix.
For LSA, the original word by document matrix is compressed to a factor by document matrix and the number of factors is much less than the number of words.
For vector-based models, each tag is represented by a compressed vector with a length much less than the number of words.

Information is lost with this compression in both cases: by starting with a compressed representation with vector-based models or by reducing to a compressed representation with LSA.
However, the way that the compression works in these two cases is not the same.
LSA uses SVD to find words that have highly correlated tag occurrence distributions (i.e., latent concepts), and then pools those co-occurrence counts into a single dimension.
A vector-based approach like random permutations essentially randomizes the mapping of the words on the reduced set of dimensions, and does not take into account word similarity when generating the mapping.
Nonetheless, vector-based models have been shown to perform similar to if not better than LSA when trained on the TASA and tested on the TOEFL \parencites{Sahlgren2008,Jones2007}.

@@

\textcite{Stanley2013} used a declarative memory model to predict a user's chosen tags for a post on the StackOverflow site.
They modified the standard ACT-R model in order to more accurately predict tag use on StackOverflow.
A formal description of the modified ACT-R Bayesian model is included in Table \ref{tabModACTRModel}.

\begin{table}[!ht]
  \caption{\citeauthor{Stanley2013}'s StackOverflow tag prediction model}
  \label{tabModACTRModel}
  {\tabulinesep=1.2mm
    \begin{tabu}{ll}
      \hline
      Common Name &  Equation \\
      \hline
      Activation & 		$A_{i} = B_{i} + \sum_{j\in T}^{ } W_{j} S_{ji} + \sum_{j\in B}^{ } W_{j} S_{ji} - O$ \\
      Base Level & 		$B_{i} = log \frac{p_{i}}{1-p_{i}}$ \\
      Strength Assoc. &		$S_{ji} = log \frac{p(i|j)}{p(i))} = log \frac{NN_{ji}}{N_{Row}(j)N_{Col}(i)}$ \\
      Attentional Weight & 	$W_{j} = W \frac{E_{j}} {\sum_{}^{} {E_{j}}} $ \\
      Scaled Entropy & 		$E_{j} = 1-\frac{H_{j}}{H_{max}}$ \\
      Entropy & 		$H_{j} = -\sum_{i=1}^{N}p(i|j)log\left (  p(i|j) \right )$ \\
      \hline
    \end{tabu}
  }
\end{table}

The activation of each tag ($A_{i}$) is a function of four components:
That tag's base-level activation ($B_{i}$), contextual activation for the words in the title, activation for words in the body, and an offset term ($O$).
This model has been modified from the standard ACT-R model in four ways:
Two separate contextual components were used (words in title and body of post), an offset term was added, an entropy measure was used to compute the attentional weight ($W_{j}$),
and the base-level activation of a tag was based on global past tag use and not customized to a user's past tagging history.

The offset term was added to normalize the mean activation across retrievals.
This allowed the use of a logistic regression statistical technique to calibrate the weights for each of the terms.
The scaled entropy measure $E_{j}$ was chosen for the attentional weight ($W_{j}$) term because it has a strong theoretical foundation \parencite{Dumais1991} and is parameter free.
Looking at $W_{j}$, if $E_{j}$ is low then $W_{j}$ is low, meaning that the scaled entropy measure helps allocate limited attentional resources to predictive contextual elements.

@@

Several hashtag recommendation models for Twitter have been developed recently.
One of the first models used a Bayesian co-occurrence statistical technique to predict the most likely hashtag associated with a post as a function of prior hashtag use and context \parencite{Mazzia2009}.
The model presented by \textcite{Mazzia2009} is quite similar to an ACT-R declarative retrieval model.
The main potential difference is that --- like \textcite{Stanley2013} --- the global prior likelihood of a hashtag is computed without taking into account the specific user's past history.

Several other models have taken a tweet-centered approach to hashtag prediction, where suggested hashtags are collected from hashtags used from similar tweets.
\textcites{Li2011, Zangerle2011, Kywe2012} all store a content vector for each tweet, and then compute a word co-occurrence-based similarity score between a composed tweet and the rest of the tweets in the database.

@@

@@

Two models were evaluated for hashtag prediction on both the Twitter and StackOverflow datasets: a vector-based model and Bayesian retrieval model.
The random permutation model \parencite{Sahlgren2008} in Table \ref{tabRandPermModel} was used for the vector-based model,
since it has shown to perform better than the original BEAGLE vector-based model \parencite{Recchia2010}.
The modified ACT-R Bayesian model \parencite{Stanley2013} in Table \ref{tabModACTRModel} was used for the Bayesian model, since it has already been applied to tag use on StackOverflow posts.

@@

An additional popular-hashtags dataset was collected for Twitter.
The purpose of this dataset is to test the accuracy of each model's predicted most likely hashtag(s) for a tweet.
The dataset contains approximately three million tweets where at least one of the 400 most popular hashtags are used.
A quota of three million tweets was selected because previously research testing Bayesian co-occurrence models on large-scale datasets has used on the order of one million documents
\parencites{Stanley2013,Douglass2010,Budiu2007}.
Tweets were collected in real time until the tweet quota was reached.

@@

A custom tokenization algorithm created by \textcite{Owoputi2013} was used to chunk the words in the text.
This algorithm was specifically created for tokenizing Twitter content.
However, during testing it was also found to be robust and well suited for StackOverflow content.
The initial model results with the new tokenizer were qualitatively similar to the model results found in \textcite{Stanley2013} where the Python Natural Language Processing (NLP) toolkit \parencite{Bird2009} was used.
So the same tokenization algorithm from \textcite{Owoputi2013} was used for both StackOverflow and Twitter, since performance was similar to \textcite{Stanley2013} for StackOverflow,
and it seemed beneficial to use the same algorithm for both datasets to make comparisons across datasets more equitable.

@@

The decay rate parameter ($d$) represents how fast the activation for an individual presentation of a chunk decays from declarative memory.
Higher values mean that chunk activation decays faster, which means that the most recently presented chunks will have the highest activation, and the total number of times that a chunk has been presented matters less.
Lower decay rate values flip the importance of frequency and recency.

@@

Both the standard form and optimized learning form of $B_{i}$ were used to predict each user's specific tag history.
To do this, a user's entire tag history was collected, and then each tag instance from oldest to newest was examined.
For each of these tagging instances, the model generated an activation value for each possible tag based on that tag's frequency and recency of past use.

@@

@@

@@

Chapter 5 will examine how model accuracy is influenced by four architectural concerns
(method of handling stop words, influence of dataset size, method to combine prior and context, and word order).
The four architectural concerns were examined for the following reasons:
During exploratory evaluation, model accuracy seemed to be highly sensitive to the method of dealing with the commonly-used stop words, so several different methods were examined in more detail.
Due to research from \textcite{Budiu2007} and \textcite{Recchia2010}, model accuracy for the Bayesian and vector-based models is influenced by the number of documents in the dataset,
so the effect of dataset size on accuracy was examined in more detail.
One of the primary goals of this research is to identify the best method to combine the model term for past user behavior with the context term for the random-permutation model,
so two different ways to add the model terms were tested.
Finally, previous research has shown that including word order in the vector-based models improves performance \parencites{Jones2007, Sahlgren2008},
so word order was also examined to see if the effect could be replicated and also to measure the overall importance of word order on model accuracy relative to other architectural manipulations.

@@

In order to best combine these predictors, a logistic regression statistical technique similar to \textcite{Stanley2013} was used to determine the most optimal weights for each term.
When a user creates a tag for a post, a retrieval request for the model is made.
For each request, the model returns a set of tags (all tags that have been observed for that user in the past) and activations associated with each tag.
An activation value is computed for each model component (e.g., prior tag use, context) for each tag.
These values are recorded, and tag instances that match what the author actually tagged are marked as a 1, while all others are marked with a 0.
This process is repeated across all posts where a user chooses a tag, and across all users in the dataset.
Once all of the results are aggregated, a logistic regression is run where the optimal weights for each model component are found that maximize the model's ability to correctly label tags as 0 or 1 based on total activation.
